Introduction

### 2020-07-30 text/discussion.Rmd

One of the biggest hurdles in cancer research is the size of the search space;
~20,000 protein-coding genes is comparable with the number of tumor samples,
even with multiple mutations per sample. And that is not accounting for the fact
that ~30k non-coding genes and hundreds of thousands of non-coding regulatory
elements exist with oncogenic or tumor suppressor activity [@deesMuSiC2012;
@tamboreroOncodriveCLUST2013;
@lawrenceDiscoverySaturationAnalysis2014;
@kumarStatisticallyIdentifyingTumor2015; @jiangWITER2019;
@zhaoModelingPositive2019; @non-coding-mutations-Zhang]. Hence, we sought to
simplify the problem by investigating cancer as a disease of basic cellular and
biochemical pathways, which we accomplished by translating gene-level mutations
into pathway level disruptions. Our approach differs from previously proposed
methods [@PATHNET] in that we chose to focus on the pathway as the unit of
disruption instead of the gene, employing the conceit that individual mutations
may be sufficient to disrupt entire pathways (as is often the case with
drivers). Rather than evaluating the burden of mutations on each pathway within
or across tumors, which can be biased by the number of genes or the coverage of
pathway-linked genes in the genome, and which we find to be relatively uniformly
distributed, we scored each pathway in binary fashion, allowing one mutation or
multiple mutations to count equally for each sample.

To our knowledge, this approach has not previously been attempted in the
literature in spite of its relative simplicity. We limited our analysis to
mutations with likely deleterious effects, including missense, nonsense,
frame-shift, stop-gain and splice site mutations in genes that are actively
expressed in each cancer type, thus avoiding bias from transcription coupled
repair. Our method of filtration differs from the "rank-and-cut" method of
[@PCAWG-MAIN] but represents a reasonable attempt to account for the same
biases. We also filtered genes on the basis of pathway membership in biochemical
pathways, excluding curated genesets related to diseases, syndromes, or classes
of proteins ( _e.g._ transcription factors). We chose this approach to limit
redundancy, which is already high in the remaining pathways for some driver
genes, and to exclude biologically unrelated collections of genes. We do not
account for non-coding mutations, amplification events or structural variants,
though we hope to include them in subsequent studies (see below). Our
pathway-centric analysis also allows us to circumvent the problem that some
cancer-promoting mutations likely occur at low enough frequencies as to be
indistinguishable from background. Previous attempts to capture these
associations better using Bayesian networks have yielded important new insights
into this problem and new genes [@hornNetSigNetworkbasedDiscovery2018] that
would not have been possible with traditional analysis, supporting the case that
pathway-centric approaches like ours have the potential to increase statistical
power.

>>>>

One of the biggest hurdles in cancer research is the size of the search space;
~20,000 protein-coding genes is comparable with the number of tumor samples,
even with multiple mutations per sample. And that is not accounting for the fact
that ~30k non-coding genes and hundreds of thousands of non-coding regulatory
elements exist with oncogenic or tumor suppressor activity [@deesMuSiC2012;
@tamboreroOncodriveCLUST2013;
@lawrenceDiscoverySaturationAnalysis2014;
@kumarStatisticallyIdentifyingTumor2015; @jiangWITER2019;
@zhaoModelingPositive2019; @noncoding-mutations-Zhang]. Hence, we sought to
simplify the problem by employing a "knowledge-base driven analysis"
[@pathways-Atul-Butte], investigating cancer as a disease of basic cellular and
biochemical pathways, which we accomplished by translating gene-level mutations
into pathway level disruptions. Our approach differs from previously described
methods [@PATHNET] in that we chose to focus on the pathway as the unit of
disruption instead of the gene, employing the conceit that individual mutations
may be sufficient to disrupt or misdirect pathway activity (as is often the case
with driver mutations). Rather than evaluating the burden of mutations on each
pathway within or across tumors, which can be biased by the number of genes or
the coverage of pathway-linked genes in the genome, we scored each pathway in
binary fashion, allowing one mutation or multiple mutations to count equally for
each sample.

To our knowledge, this approach has not previously been attempted in the
literature in spite of its relative simplicity. We limited our analysis to
mutations with likely deleterious effects, including missense, nonsense,
frame-shift, stop-gain and splice site mutations in genes that are actively
expressed in each cancer type, thus avoiding bias from transcription coupled
repair. Our method of filtration differs from the "rank-and-cut" method of
[@PCAWG-MAIN] but represents a reasonable attempt to account for the same
biases. We also filtered genes on the basis of pathway membership in biochemical
pathways, excluding curated genesets related to diseases, syndromes, or classes
of proteins ( _e.g._ transcription factors) which are potentially problematic
[@pathways-Atul-Butte]. We chose this approach to limit redundancy, which is
already high in the remaining pathways for some driver genes, and to exclude
biologically unrelated collections of genes. We do not account for non-coding
mutations, amplification events or structural variants, though we hope to
include them in subsequent studies (see below). Our pathway-centric analysis
also allows us to circumvent the problem that some cancer-promoting mutations
likely occur at low enough frequencies as to be indistinguishable from
background. Previous attempts to capture these associations better using
Bayesian networks have yielded important new insights into this problem and new
genes [@hornNetSigNetworkbasedDiscovery2018] that would not have been possible
with traditional analysis, supporting the case that pathway-centric approaches
like ours have the potential to increase statistical power.

### 2020-07-30 text/introduction.Rmd
add temporary header to discussion and intro to compile references, fix typos, update discussion edits

@@

More recently, attempts have been made to consider pan-cancer mutations and
other data outside the context of enrichment at the gene level. Horn _et al._
[@hornNetSigNetworkbasedDiscovery2018] introduce approaches to address the
sparseness of mutations, allowing the potential importance of a gene to be
influenced by the aggregate of positive selection in its nearest network
neighbors. This method is powerful especially when applied to a small number (on
the order of hundreds) of tumors. However, the larger problem of cancer etiology
remains: what mutations are required, in what order, and are there specific
combinations of delinquent cellular processes sufficient to cause cancer? If
there is a common logic to cancer, how is it to be detected underneath the
overwhelming cell-of-origin signal?

In this study, we attempt to address these questions by focusing on discrete
molecular pathways as the unit of disruption in cancer. Our approach strips all
cell-type-specific information from the mutation data and generalizing
gene-level information to non-tissue specific cell-biological pathway
phenotypes. We use this rubric to evaluate all cancers and show that,
surprisingly, tumors may have distinct phenotypes completely independent of
their tissues of origin.

>>>>

---
title: "Introduction"
output: 
  bookdown::html_document2:
    fig_captions: TRUE
    number_sections: FALSE
    toc: FALSE
    theme: flatly
bibliography: references.bib
---

@@

More recently, attempts have been made to consider pan-cancer mutations and
other data outside the context of enrichment at the gene level. Horn _et al._
[@hornNetSigNetworkbasedDiscovery2018] introduce approaches to address the
sparseness of mutations, allowing the potential importance of a gene to be
influenced by the aggregate of mutations in its nearest network
neighbors. This method is powerful especially when applied to a small number (on
the order of hundreds) of tumors. However, the larger problem of cancer etiology
remains: what mutations are required, in what order, and are there specific
combinations of delinquent cellular processes sufficient to cause cancer? If
there is a common logic to cancer, how is it to be detected underneath the
overwhelming cell-of-origin signal?

In this study, we attempt to address these questions by focusing on discrete
molecular pathways as the unit of disruption in cancer. Our approach strips all
cell-type-specific information from the mutation data and generalizing
gene-level information to non-tissue specific cell-biological pathway
phenotypes. We use this statistic to evaluate all cancers and show that,
surprisingly, tumors may have distinct phenotypes completely independent of
their tissues of origin.
